\chapter*{Acknowledgments}

First and foremost obviously I wish to thank my advisors and mentors Sergi Jordà and Perfecto Herrera. Thanks to Sergi for meeting me in Belfast many moons ago and bringing me to Barcelona. I thank him for his constant commitment, dedication and above all, inspiration. I am equally thankful to Perfecto Herrera for his vast knowledge, infinite patience and unwavering support. 

To the most gigantic steppers of them all, the musketeers: Ángel, Dani and Martin, one for all and all for one.

Thanks to Xavier, Cristina, Sonia, Lydia and the rest of UPF for making it a great environment for this lowly scholar.

Shout out to all the MTG gang: Juanjo (particularly for the translations, squawk), Marius, Alastair, Sergio, Frederick, Dimi, Oriol, Moha, Rafa, Sebas, Carles, Georgi. Dara, Gerard, Álvaro, Pano, Giuseppe, Martí etc. I am missing others I know but know that it is not intentional. 

Cheers to the Barcelona Laptop Orchestra for their collaboration.

Thanks to the Barcelona crew for keeping me (in)sane these past years, and the Bonitos for keeping the mandolin strings from rusting.

Thanks to my Cork-based friends and my Dublin-based friends.

I am indebted to Peter for reviewing some drafts.

Faoi dheireadh caithfidh an buíochas is mó mé thabhairt  do mo chlann, as ocht an grá agus an  tachaíocht a thug sibh dom i rith an bhliain dorcha, deachair seo a bhí againn go léir. Go raibh maith agat James \& Batchimeg, Colm \& Melanie 'gus fáilte go dtí an domhan Marló beagaín! Míle buíochas freisin go dtí muintir McCarthy agus Ó Nuanáin.

Go raibh maith agat a thuismitheoirí, ar dtús báire as ocht thabhairt dom an cheoil, mar ní bheadh mé abálta é seo a dhéanamh gan é. Go raibh maith agat, Joe as gach rud, ach go háirithe an cabhair, an bia, na deochanna, agus an Béarla!

Agus le sin, go raibh míle maith agat Marian, tá brón orm níl tú in ann seo a léamh a chroí.
%
%\vspace*{\fill}
%
%\line(1,0){372}\\
%\footnotesize
%This thesis has been carried out at the Music Technology Group of Universitat Pompeu Fabra (UPF) in Barcelona, Spain, from Oct.~2012 to Sep.~2016. This work has been supported by the Dept. of Information and Communication Technologies (DTIC) PhD fellowship (2012-16), Universitat Pompeu Fabra and the European Research Council under the European Union’s Seventh Framework Program, as part of the CompMusic project (ERC grant agreement 267583).
\normalsize